Optimizing Abstract Abstract Machines

### 2012-11-12 paper/paper.tex
Talked about mutable literal data, gave a breakdown of the implementation (in comments) and gave better explanation for deltas.

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}

@@

In section~\ref{sec:impl}, we describe a more complete implementation
scaled up and capable of analyzing Scheme benchmark programs.

In section~\ref{sec:eval} we evaluate this implementation against a
set of benchmark programs drawn from the literature.  Many benchmarks
cause the baseline analyzer to take longer than 30 minutes or to
consume more 1 gigabit of memory, at which point the analysis is
stopped.  This is the case for the largest benchmark program, which is
3,500 lines of code and takes under a minute in the most optimized
analyzer.  For those benchmarks that did complete on the baseline, the
optimized analyzer outperformed the baseline by a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector (preallocation)}
 \item{[SS] coarsely abstract literal data (abstracting compound data)}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors. The first three and last remain in the purely
functional realm. The final three use mutation in small, localized
places, so they do not increase conceptual complexity
drastically. Also, preallocation and mutable workset are standard
implementation practices \cite{ianjohnson:Shivers:1991:CFA}.

@@

Laziness removes several represented states that would then have to be
stepped by the fixpoint combinator.  We can further eliminate the {\tt
  ev} states from the run-time interpretation of a program by
specializing the machine transition relation to the program being
analyzed. This eliminates interpretative overhead by first compiling
the program into ``byte code'' instructions.

@@

\begin{figure}
\begin{align*}
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\label{fig:compile}
\end{figure}

@@

Each step in the fixpoint for the above techniques requires joining
several, commonly large, stores together. Even with immutable hashes
that utilize sharing, this is memory- and time-intensive. Worse, this
contributes an obscene fraction of the runtime. If we instead track
the changes to the store made over all steps and afterward update the
store, we end up with much less allocation and computational overhead.

We represent changes as a list of pairs of address to set of storeable
values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
each step. Applying the changes is straightforward:

@@

\begin{align*}
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}
\begin{align*}
bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr'
\end{align*}

We also lift $\machstep$ to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated.

@@

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before (vis. $f(X)
\setminus S$ in ${\mathcal G}$), we still have to compare entire stores - an
expensive computation.  We modify a technique created in Shivers'
dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
approximation, that removes the need for joining whole stores by
manipulating a global store imperatively. The original technique loses
precision because the starting store for each state in a machine
configuration can change between stepping all states and introduce
more flows than originally possible. However, by delaying the store
changes until after all steps have completed, we can maintain our
precision. In a variety of experiments, we observed that the overhead
for storing store-deltas was outweighed by the precision keeping flows
in check. That is to say, the analysis was faster {\it and} more
precise.

We pair this technique with a mutable workset. That is, to avoid the
intermediate sets introduced by the $State \to \Set(State)$ signature
of $\machstep$, we change the fixpoint algorithm to use a global
workset, {\tt todo}, that $\machstep$ imperatively updates with the
states that it reaches.

In total, we have six globals:
\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from state to the timestamp at which it was last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

@@

\subsection{Abstracting Compound Data}

@@

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is dramatically smaller in typical Scheme programs. We
use the above abstraction for these syntactic uses, but choose to
interpret literal lists as not always sugar for cascading {\tt
  cons}es. In particular, if a list literal is ``too big'' (in our
case length > 1), we interpret the list as a circular data structure;
the right address points back to the {\tt cons} itself, and the left
address points to all of the elements of the list. We take this
farther and join the list elements together in with a coarse
type-based abstraction. Quotation is special because it cannot
introduce function values, which is important to enhancing our
technique's soundness to conceptual complexity ratio. If we join two
values of different type together, we don't get ``anything,'' which
has complex meaning in higher-order languages (Shivers' escape
semantics) and is overly approximate. We instead get ``any quotable
value,'' which has much simpler semantics.

@@

@@

@@

\bibliographystyle{plain}
\bibliography{local,bibliography}

>>>>

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}
\usepackage{balance}

@@

In section~\ref{sec:eval} we evaluate our implementation against a
set of benchmark programs drawn from the literature.  Many benchmarks
cause the baseline analyzer to take longer than 30 minutes or to
consume more 1 gigabit of memory, at which point the analysis is
stopped.  This is the case for the largest benchmark program, which is
3,500 lines of code and takes under a minute in the most optimized
analyzer.  For those benchmarks that did complete on the baseline, the
optimized analyzer outperformed the baseline by a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

\begin{itemize}
 \item{[SS] lazy non-determinism: quotient by sets of values}
 \item{[SS] abstract compilation: eliminate {\tt ev} states}
 \item{[DS] store deltas: operate on store changes rather than whole stores}
 \item{[DS] timestamp: represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting}
 \item{[DS] preallocation: represent the global store as a global, mutable vector}
 \item{[SS] uniform literal approximation: coarsely abstract literal data}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors.

@@

Laziness removes several represented states that would then have to be
stepped by the fixpoint combinator.  We can further eliminate the {\tt
  ev} states from the run-time interpretation of a program by
specializing the machine transition relation to the program being
analyzed. This precision-preserving transformation eliminates
interpretative overhead by first compiling the program into ``byte
code'' instructions.

@@

\begin{figure}
\begin{align*}
\compile{\_} &: Expr \to Env \times Store \times Kont \to State \\
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\label{fig:compile}
\end{figure}

@@

Each step in the fixpoint for the above techniques requires joining
several, commonly large, stores together. Each step is unlikely to
always modify all addresses of the store, so this is needlessly
memory- and time-intensive. We can instead accumulate store changes
and apply them after all steps have completed. This uses far fewer
join operations, leading to much less allocation and computational
overhead.

We represent changes as $Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
begins as empty for each step. Applying the changes is
straightforward:

@@

\begin{align*}
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}
\begin{align*}
bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr'
\end{align*}
Compilation changes to additionally take a $\msdiff$ component, so the
above rule's right hand side would instead be $k^{\mcntr'}(\menv',
\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.

We also lift $\machstep$ to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated:

@@

The store delta technique removes overhead of joining large stores,
but for determining if a state has been visited before (vis. $f(X)
\setminus S$ in ${\mathcal G}$), we still have to compare entire
stores - an expensive computation.  We modify a technique due to
Shivers~\cite{ianjohnson:Shivers:1991:CFA}, timestamp approximation,
that removes the need for joining whole stores by manipulating a
global store imperatively. The original technique loses precision
because states are stepped in sequence, modifying the store in place -
more values or continuations can be added before another state gets to
step. This interference means more possible flows (i.e. less
precision) and a nondeterministic analysis - the order the set of states is
traversed is not necessarily deterministic. However, by delaying the
store changes until after all steps have completed, we can maintain
our precision and determinism. In a variety of experiments, we
observed that the time and space overhead for storing store-deltas was outweighed by
the precision keeping flows in check. That is to say, the analysis was
faster {\it and} more precise.

We pair this technique with a mutable workset to avoid the
intermediate sets introduced by the $State \times Store \to \Set(State) \times Store\msdiff$ signature
of $\machstep$. This necessitates a change to the fixpoint algorithm to use a global
workset, {\tt todo}, that $\machstep$ imperatively updates with the
states that it would have otherwise accumulated in a set that would later be added to the workset.

In total, we have six globals:
\begin{itemize}
 \item{{\tt $\msto$}, the store}
 \item{{\tt todo}, the workset}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
 \item{{\tt $\msdiff$}, the store changes for the current step}
 \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
 \item{{\tt T}, the timestamp of the store}
\end{itemize}

@@

\subsection{Abstracting literal compound data}

@@

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is dramatically smaller in typical Scheme programs. We
use the above abstraction for these syntactic uses, but choose to
interpret literal lists as not always sugar for cascading {\tt
  cons}es. In particular, if a list literal is ``too big'' (in our
case length $>$ 1), we interpret the list as a circular data structure;
the right address points back to the {\tt cons} itself, and the left
address points to all of the elements of the list. We take this
farther and join the list elements together in with a coarse
type-based abstraction. Quotation is special because it cannot
introduce function values, which is important to enhancing our
technique's soundness to conceptual complexity ratio. If we join two
values of different type together, we don't get ``anything,'' which
has complex meaning in higher-order languages (Shivers' escape
semantics) and is overly approximate. We instead get ``any quotable
value,'' which has much simpler semantics.

@@

\paragraph{Mutable literal data} The Scheme we model is nonstandard in that it has immutable literal
data (the Scheme benchmarks did not mutate their literal data). To
support mutable literal data, abstract values like {\tt QPair} would
be like regular conses, only point to a preallocated pair of addresses
that hold {\tt QData} as well. This is because conses can be mutated
to hold any value, not just quoted data.

@@

Our implementation of the analysis used in the evaluation is roughly
3500 lines of Racket code.
%% 400 lines for the core, 700 lines for
%% abstraction over optimizations, 1500 lines for primitives and standard
%% list operations, 700 lines for instantiations to different
%% optimizations and 300 for parsing and macro transformers.

@@

\balance
\bibliographystyle{plain}
\bibliography{local,bibliography}
